\bf Calendar-based graphics for visualizing people's daily schedules

### 2018-10-20 main.Rmd
many tweaks

The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area (Figure \ref{fig:ped-map}). The dataset can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. A routine examination of the data would involve constructing conventional time series plots to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at three different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse of the daily and sub-daily pedestrian patterns.

@@

However, the conventional displays of time series data obscure patterns relative to special events (such as public holidays and recurring cultural/sport events), which may appear more fascinating to viewers.

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plots showing the number of pedestrians in 2016 measured at three different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these three sensors. There are eye-catching spikes which occurred at the State Library and Birrarung Marr, caused by the annual White Night event on 20th of February."}
```

```{r facet-time, fig.height = 6, fig.cap = "Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library."}
```

@@

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:opt} lists and describes the options that come with the `frame_calendar` function. Section \ref{sec:variations} presents some variations of its usage. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses the limitations of calendar displays and possible new directions.

@@

## Data transformation

@@

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", 
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, 
  scale = "fixed", width = 0.95, height = 0.95, margin = NULL
)
```

@@

When `polar = TRUE`, a polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:fs-polar} shows star plots embedded in the monthly calendar layout, which is equivalent to Figure \ref{fig:fs-2016} placed in polar coordinates. The bimodal workday shape is also visible as boomerangs.

@@

```{r overlay, fig.height = 7, fig.cap = "Overlaying line graphs of the three sensors in the monthly calendar. Three sensors demonstrate very different traffic patterns. Birrarung Marr tends to attract many pedestrians for special events held on weekends, by contrasting to the bimodal commuting traffic at Flagstaff Station."}
```

To avoid the overlapping problem, the calendar layout can be embedded into a series of subplots for the different sensors. Figure \ref{fig:facet} presents the idea of faceting calendar plots. This allows comparing the overall structure between sensors, while emphasising individual sensor variation.

```{r facet, fig.height = 10, fig.cap = "Line charts, embedded in the $6 \\times 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors."}
```

@@

As a data re-structuring tool, the interactivity of calendar-based display can be easily enabled, as long as the interactive graphic system remains true to the spirit of the grammar of graphics, for example **plotly** in R. As a standalone display, interactive tooltip can be added to show the actual values of a data point when mousing over it in the calendar layout, for example the hourly count with the time of day. It is difficult to sense the values in static display, but the tooltip makes it possible. Options in function `frame_calendar` can be ported to a form of selection buttons or text input. The display will update on the fly accordingly via clicking or inputing, as desired.

@@

The use of the calendar display is illustrated on smart meter energy usage from four households in Melbourne, Australia. Individuals can download their own data from the energy company, and these four households are the data of colleagues of the authors. The calendar display is useful to help people understand their energy use. The data contains half-hourly electricity consumption in 2017 and 2018. The analysis begins by looking at the distribution over days of week, then time of day split by workdays and non-work days, followed by the calendar display to inspect the daily schedules.

>>>>

The purpose of the calendar-based visualization is to provide insights into people's daily schedules relative to events such as work days, weekends, holidays, and special events. This work was originally motivated by studying foot traffic in the city of Melbourne, Australia [@ped]. There have been 43 sensors installed that count pedestrians every hour across the inner-city area till the end of 2016 (Figure \ref{fig:ped-map}). The dataset can shed light on people's daily rhythms, and assist the city administration and local businesses with event planning and operational management. A routine examination of the data would involve constructing conventional time series plots to catch a glimpse of temporal patterns. The faceted plots in Figure \ref{fig:time-series-plot}, give an overall picture of the foot traffic at three different sensors over 2016. Further faceting by day of the week (Figure \ref{fig:facet-time}) provides a better glimpse of the daily and sub-daily pedestrian patterns.

@@

However, the conventional displays of time series data obscure patterns relative to special events (such as public holidays and recurring cultural/sport events), which may appear fascinating to viewers.

@@

```{r time-series-plot, fig.height = 6, fig.cap = "Time series plots showing the number of pedestrians in 2016 measured at three different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. A spike occurred at the State Library, caused by the annual White Night event on 20th of February. A relatively fixed pattern repeats from one week to another at Flagstaff Station. Birrarung Marr looks rather random and spiky, with chunks of missing records."}
```

```{r facet-time, fig.height = 6, fig.cap = "Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It primarily features two types of seasons---time of day and day of week---across all the sensors. Apparently other factors have influence over the number of pedestrians, which cannot captured by the faceted plots, such as the overnight White Night traffic on Saturday at the State Library and a variety of events at Birrarung Marr."}
```

@@

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:transformation} describes the algorithms of data transformation. Section \ref{sec:opt} lists and describes the options that come with the `frame_calendar` function. Section \ref{sec:variations} presents some variations of its usage. Graphical analyses of sub-daily people's activities are illustrated with a case study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses the limitations of calendar displays and possible new directions.

@@

## Data transformation
\label{sec:transformation}

@@

```
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h", 
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed", 
  width = 0.95, height = 0.95, margin = NULL)
```

@@

When `polar = TRUE`, a polar transformation is carried out on the data. The computation is similar to the one described in @Wickham2012glyph. Figure \ref{fig:fs-polar} shows star plots embedded in the monthly calendar layout, which is equivalent to Figure \ref{fig:fs-2016} placed in polar coordinates. The bimodal work day shape is also visible as boomerangs.

@@

```{r overlay, fig.height = 7, fig.cap = "Overlaying line graphs of the three sensors in the monthly calendar. Three sensors demonstrate very different traffic patterns. Birrarung Marr tends to attract many pedestrians for special events held on weekends, contrasting to the bimodal commuting traffic at Flagstaff Station."}
```

To avoid the overlapping problem, the calendar layout can be embedded into a series of subplots for the different sensors. Figure \ref{fig:facet} presents the idea of faceting calendar plots. This allows comparing the overall structure between sensors, while emphasising individual sensor variation. In particular, we can immediately learn that when Birrarung Marr was crowded and why, such as Australian Open in the last two weeks of January. This does not pop out in the conventional graphics.

```{r facet, fig.height = 10.2, fig.cap = "Line charts, embedded in the $6 \\times 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors."}
```

@@

As a data restructuring tool, the interactivity of calendar-based display can be easily enabled, as long as the interactive graphic system remains true to the spirit of the grammar of graphics, for example **plotly** [@plotly] in R. As a standalone display, interactive tooltip can be added to show the actual values of a data point when mousing over it in the calendar layout, for example the hourly count with the time of day. It is difficult to sense the values in static display, but the tooltip makes it possible. Options in function `frame_calendar` can be ported to a form of selection buttons or text input. The display will update on the fly accordingly via clicking or inputing, as desired.

@@

The use of the calendar display is illustrated on smart meter energy usage from four households in Melbourne, Australia. Individuals can download their own data from the energy company, and these four households are the data of colleagues of the authors. The calendar display is useful to help people understand their energy use. The data contains half-hourly electricity consumption in 2017 and 2018. The analysis begins by looking at the distribution over days of week, then time of day split by work days and non-work days, followed by the calendar display to inspect the daily schedules.

### 2018-10-20 main.tex
many tweaks

The purpose of the calendar-based visualization is to provide insights
into people's daily schedules relative to events such as work days,
weekends, holidays, and special events. This work was originally
motivated by studying foot traffic in the city of Melbourne, Australia
\citep{ped}. There have been 43 sensors installed that count pedestrians
every hour across the downtown area (Figure \ref{fig:ped-map}). The
dataset can shed light on people's daily rhythms, and assist the city
administration and local businesses with event planning and operational
management. A routine examination of the data would involve constructing
conventional time series plots to catch a glimpse of temporal patterns.
The faceted plots in Figure \ref{fig:time-series-plot}, give an overall
picture of the foot traffic at three different sensors over 2016.
Further faceting by day of the week (Figure \ref{fig:facet-time})
provides a better glimpse of the daily and sub-daily pedestrian
patterns.

However, the conventional displays of time series data obscure patterns
relative to special events (such as public holidays and recurring
cultural/sport events), which may appear more fascinating to viewers.

@@

\caption{Time series plots showing the number of pedestrians in 2016 measured at three different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these three sensors. There are eye-catching spikes which occurred at the State Library and Birrarung Marr, caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}

@@

\caption{Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library.}\label{fig:facet-time}
\end{figure}

@@

The remainder of the paper is organized as follows. Section
\ref{sec:algorithm} demonstrates the construction of the calendar layout
in depth. Section \ref{sec:opt} lists and describes the options that
come with the \texttt{frame\_calendar} function. Section
\ref{sec:variations} presents some variations of its usage. Graphical
analyses of sub-daily people's activities are illustrated with a case
study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses
the limitations of calendar displays and possible new directions.

@@

@@

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", 
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, 
  scale = "fixed", width = 0.95, height = 0.95, margin = NULL
)
\end{verbatim}

@@

When \texttt{polar\ =\ TRUE}, a polar transformation is carried out on
the data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:fs-polar} shows star plots
embedded in the monthly calendar layout, which is equivalent to Figure
\ref{fig:fs-2016} placed in polar coordinates. The bimodal workday shape
is also visible as boomerangs.

@@

\caption{Overlaying line graphs of the three sensors in the monthly calendar. Three sensors demonstrate very different traffic patterns. Birrarung Marr tends to attract many pedestrians for special events held on weekends, by contrasting to the bimodal commuting traffic at Flagstaff Station.}\label{fig:overlay}
\end{figure}

To avoid the overlapping problem, the calendar layout can be embedded
into a series of subplots for the different sensors. Figure
\ref{fig:facet} presents the idea of faceting calendar plots. This
allows comparing the overall structure between sensors, while
emphasising individual sensor variation.

@@

As a data re-structuring tool, the interactivity of calendar-based
display can be easily enabled, as long as the interactive graphic system
remains true to the spirit of the grammar of graphics, for example
\textbf{plotly} in R. As a standalone display, interactive tooltip can
be added to show the actual values of a data point when mousing over it
in the calendar layout, for example the hourly count with the time of
day. It is difficult to sense the values in static display, but the
tooltip makes it possible. Options in function \texttt{frame\_calendar}
can be ported to a form of selection buttons or text input. The display
will update on the fly accordingly via clicking or inputing, as desired.

>>>>

The purpose of the calendar-based visualization is to provide insights
into people's daily schedules relative to events such as work days,
weekends, holidays, and special events. This work was originally
motivated by studying foot traffic in the city of Melbourne, Australia
\citep{ped}. There have been 43 sensors installed that count pedestrians
every hour across the inner-city area till the end of 2016 (Figure
\ref{fig:ped-map}). The dataset can shed light on people's daily
rhythms, and assist the city administration and local businesses with
event planning and operational management. A routine examination of the
data would involve constructing conventional time series plots to catch
a glimpse of temporal patterns. The faceted plots in Figure
\ref{fig:time-series-plot}, give an overall picture of the foot traffic
at three different sensors over 2016. Further faceting by day of the
week (Figure \ref{fig:facet-time}) provides a better glimpse of the
daily and sub-daily pedestrian patterns.

However, the conventional displays of time series data obscure patterns
relative to special events (such as public holidays and recurring
cultural/sport events), which may appear fascinating to viewers.

@@

\caption{Time series plots showing the number of pedestrians in 2016 measured at three different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. A spike occurred at the State Library, caused by the annual White Night event on 20th of February. A relatively fixed pattern repeats from one week to another at Flagstaff Station. Birrarung Marr looks rather random and spiky, with chunks of missing records.}\label{fig:time-series-plot}
\end{figure}

@@

\caption{Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It primarily features two types of seasons---time of day and day of week---across all the sensors. Apparently other factors have influence over the number of pedestrians, which cannot captured by the faceted plots, such as the overnight White Night traffic on Saturday at the State Library and a variety of events at Birrarung Marr.}\label{fig:facet-time}
\end{figure}

@@

The remainder of the paper is organized as follows. Section
\ref{sec:algorithm} demonstrates the construction of the calendar layout
in depth. Section \ref{sec:transformation} describes the algorithms of
data transformation. Section \ref{sec:opt} lists and describes the
options that come with the \texttt{frame\_calendar} function. Section
\ref{sec:variations} presents some variations of its usage. Graphical
analyses of sub-daily people's activities are illustrated with a case
study in Section \ref{sec:case}. Section \ref{sec:discussion} discusses
the limitations of calendar displays and possible new directions.

@@

\label{sec:transformation}

@@

\begin{verbatim}
frame_calendar(data, x, y, date, calendar = "monthly", dir = "h", 
  sunday = FALSE, nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed", 
  width = 0.95, height = 0.95, margin = NULL)
\end{verbatim}

@@

When \texttt{polar\ =\ TRUE}, a polar transformation is carried out on
the data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:fs-polar} shows star plots
embedded in the monthly calendar layout, which is equivalent to Figure
\ref{fig:fs-2016} placed in polar coordinates. The bimodal work day
shape is also visible as boomerangs.

@@

\caption{Overlaying line graphs of the three sensors in the monthly calendar. Three sensors demonstrate very different traffic patterns. Birrarung Marr tends to attract many pedestrians for special events held on weekends, contrasting to the bimodal commuting traffic at Flagstaff Station.}\label{fig:overlay}
\end{figure}

To avoid the overlapping problem, the calendar layout can be embedded
into a series of subplots for the different sensors. Figure
\ref{fig:facet} presents the idea of faceting calendar plots. This
allows comparing the overall structure between sensors, while
emphasising individual sensor variation. In particular, we can
immediately learn that when Birrarung Marr was crowded and why, such as
Australian Open in the last two weeks of January. This does not pop out
in the conventional graphics.

@@

As a data restructuring tool, the interactivity of calendar-based
display can be easily enabled, as long as the interactive graphic system
remains true to the spirit of the grammar of graphics, for example
\textbf{plotly} \citep{plotly} in R. As a standalone display,
interactive tooltip can be added to show the actual values of a data
point when mousing over it in the calendar layout, for example the
hourly count with the time of day. It is difficult to sense the values
in static display, but the tooltip makes it possible. Options in
function \texttt{frame\_calendar} can be ported to a form of selection
buttons or text input. The display will update on the fly accordingly
via clicking or inputing, as desired.
